---
title: "miniprojectR"
author: "Rachel Kraft"
date: "2/13/2022"
output: pdf_document
---

> Download and import data set

```{r}
fna.data <- "C:/Users/Rachel/Desktop/WisconsinCancer.csv"
wisc.df <- read.csv(fna.data, row.names=1)

```


> Remove diagnosis column

```{r}
wisc.data <- wisc.df[,-1]
```

> Vector for info from diagnosis column

Make a factor for diagnosis to assign the possible levels to the diagnosis column

```{r}
diagnosis_levels <- c("B", "M")
diagnosis <- factor(c(wisc.df[,1]), levels=diagnosis_levels)
```

> Q1) How many observations are in this dataset?

```{r}
dim(wisc.data)
```
> Q2) How many of the observations have malignant diagnosis?

```{r}
## grep() search for character match in a character vector
length(grep("M", diagnosis))
```

> Q3) How many variables/features in the data are suffixed with _mean?

- look for which columns have mean in them

```{r}
length(grep("mean", colnames(wisc.data)))
```

> PCA

- check mean and stdev to see if it needs to be scaled

```{r}
colMeans(wisc.data)
apply(wisc.data,2,sd)
```

```{r}
wisc.pr <- prcomp(wisc.data, scale=TRUE)
summary(wisc.pr)
```


> Q4) From your results, what proportion of the original variance is captured by the first principal components (PC1)?

When you look at the PCA results above, the proportion of variance for PC1 is 0.4427

> Q5) How many principal components (PCs) are required to describe at least 70% of the original variance in the data?

To find this, look at the cumulative proportion row for each PC column above. The first one over 0.7 is PC3, so 3 principal components are required to describe at least 70% of the original variance

> Q6) How many principal components (PCs) are required to describe at least 90% of the original variance in the data?

Cumulative proportion is at least 0.9 at PC7, so 7 principal components are required to describe at least 90% of original variance.

> Make a biplot for our PCA results

```{r}
biplot(wisc.pr)
```

> Q7) What stands out to you about this plot? Is it easy or difficult to understand? Why?

It is too cluttered and chaotic, very difficult to understand.

> Make scatter plot for each observation along PC1 and 2 

```{r}
plot(wisc.pr$x[,1:2], col=diagnosis, xlab= "PC1", ylab="PC2")
```

> Q8) Generate a similar plot for principal components 1 and 3. What do you notice about these plots?

```{r}
plot(wisc.pr$x[,1], wisc.pr$x[,3], col=diagnosis, xlab="PC1", ylab="PC3")
```

> Use ggplot2 for a better figure

```{r}
df <- as.data.frame(wisc.pr$x)
df$diagnosis <- diagnosis

library(ggplot2)

ggplot(df)+aes(PC1, PC2, col=df$diagnosis) + geom_point()
```

> Calculate variance of each PC by squaring sdev

```{r}
pr.var <- wisc.pr$sdev^2
head(pr.var)
```

> Calculate variance explained by each PC and plot

```{r}
pve <- pr.var/sum(pr.var)

plot(pve, xlab = "Principal Component", 
     ylab = "Proportion of Variance Explained", 
     ylim = c(0, 1), type = "o")
```

> Alternative plot

```{r}
barplot(pve, ylab = "Precent of Variance Explained",
     names.arg=paste0("PC",1:length(pve)), las=2, axes = FALSE)
axis(2, at=pve, labels=round(pve,2)*100 )
```

> Q9) For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean?

```{r}
wisc.pr$rotation[,1]
```

component for concave.points_mean is -0.26085376 

> Q10) What is the minimum number of principal components required to explain 80% of the variance of the data?

5 principal components are required to explain 80% of variance of the data according to the PCA analysis

> Hierarchical clustering

- scale the data

```{r}
data.scaled <- scale(wisc.data)
```

- calculate distances

```{r}
data.dist <- dist(data.scaled)
```

- create clustering model

```{r}
wisc.hclust <- hclust(data.dist, method="complete")
```

> Q11) Using the plot() and abline() functions, what is the height at which the clustering model has 4 clusters?

```{r}
plot(wisc.hclust)
abline(h=19, col="red", lty=2)
```

- cut into 4 clusters

```{r}
wisc.hclust.clusters <- cutree(wisc.hclust, k=4)
table(wisc.hclust.clusters, diagnosis)
```

> Q12) Can you find a better cluster vs diagnoses match by cutting into a different number of clusters between 2 and 10?

```{r}
for (i in c(2:10)) {
  wisc.hclust.clusters <- cutree(wisc.hclust, k=i)
  table(wisc.hclust.clusters, diagnosis)
}
wisc.hclust.clusters <- cutree(wisc.hclust, k=4)
```



Cutting into 5 clusters may be a good match too

> Q13) Which method gives your favorite results for the same data.dist dataset? Explain your reasoning.

```{r}
wisc.hclust <- hclust(data.dist, method="single")
plot(wisc.hclust)
```

```{r}
wisc.hclust <- hclust(data.dist, method="average")
plot(wisc.hclust)
```

```{r}
wisc.hclust <- hclust(data.dist, method="ward.D2")
plot(wisc.hclust)
```

I like the results from method="Ward.D2" because it looks the cleanest and allows me to see the clusters more clearly.

> Use min PA to describe at least 90% of variability and method="ward.D2"

```{r}
wisc.pr.hclust <- hclust(dist(wisc.pr$x[,1:7]), method="ward.D2")
plot(wisc.pr.hclust)
```

> Are these 2 clusters malignant and benign?

```{r}
grps <- cutree(wisc.pr.hclust, k=2)
table(grps)
```

```{r}
table(grps, diagnosis)
```

```{r}
plot(wisc.pr$x[,1:2], col=grps)
```

```{r}
plot(wisc.pr$x[,1:2], col=diagnosis)
```

```{r}
g <- as.factor(grps)
levels(g)
```

```{r}
g <- relevel(g,2)
levels(g)
```

```{r}
# Plot using our re-ordered factor 
plot(wisc.pr$x[,1:2], col=g)
```

```{r}
## Use the distance along the first 7 PCs for clustering i.e. wisc.pr$x[, 1:7]

wisc.pr.hclust <- hclust(dist(wisc.pr$x[,1:7]), method="ward.D2")

## Cut this hierarchical clustering model into 2 clusters and assign the results to wisc.pr.hclust.clusters.

wisc.pr.hclust.clusters <- cutree(wisc.pr.hclust, k=2)
table(wisc.pr.hclust.clusters, diagnosis)

```

> Q15) How well does the newly created model with four clusters separate out the two diagnoses?

This method separates the two diagnoses by majority, but i think could be better

> Q16) How well do the k-means and hierarchical clustering models you created in previous sections (i.e. before PCA) do in terms of separating the diagnoses? 

It seems that the k-means clustering model from the example did a better job at separating the diagnoses because there is a higher proportion of malignant samples in the cluster where malignant is the majority than out hierarchical clustering model. But, the hierarchical model still separated them pretty well.

```{r}
data.scaled <- scale(wisc.data)
data.dist <- dist(data.scaled)
wisc.hclust <- hclust(data.dist, method="complete")
wisc.hclust.clusters <- cutree(wisc.hclust, k=4)
table(wisc.hclust.clusters, diagnosis)
```


> Q17)  Which of your analysis procedures resulted in a clustering model with the best specificity? How about sensitivity?

-sensitivity

wisc.km$cluster

```{r}
175/(175+37)
```
wisc.hclust.clusters

```{r}
165/(165+5+40+2)
```
The wisc.km$cluster analysis has a higher sensitivity

-specificity 

wisc.km$cluster

```{r}
343/(343+14)
```
wisc.hclust.clusters

```{r}
343/(12+2+343)
```
The sensitivity of the two analyses is the same

> Prediction

```{r}
#url <- "new_samples.csv"
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc
```
```{r}
plot(wisc.pr$x[,1:2], col=g)
points(npc[,1], npc[,2], col="blue", pch=16, cex=3)
text(npc[,1], npc[,2], c(1,2), col="white")
```

> Q18) Which of these new patients should we prioritize for follow up based on your results?

We should follow up on patient 2