---
title: "Lab14: COVID-19 vaccination rate mini project"
author: "Rachel Kraft"
date: "2022-03-03"
output: 
  pdf_document: 
    latex_engine: xelatex
---


read data set for vaccination rates

```{r}
vax <- read.csv("covid19vaccinesbyzipcode_test.csv")
head(vax)
```

> Q1. What column details the total number of people fully vaccinated? 

persons_fully_vaccinated

> Q2. What column details the Zip code tabulation area? 

zip_code_tabulation_area

> Q3. What is the earliest date in this dataset? 

2021-01-05

> Q4. What is the latest date in this dataset?

```{r}
vax$as_of_date[nrow(vax)]
```

2022-03-01

> call the skim() function from the skimr package to get a quick overview of this dataset

```{r, message=FALSE}
#install.packages("skimr")
library(skimr)
```

```{r}
skim(vax)
```


> Q5. How many numeric columns are in this dataset? 

9

> Q6. Note that there are “missing values” in the dataset. How many NA values there in the persons_fully_vaccinated column?

18338

> Q7. What percent of persons_fully_vaccinated values are missing (to 2 significant figures)?

```{r}
round((18338/107604*100),2)
```


> Working with Dates

we can use the lubridate package to more easily work with dates

```{r, message=FALSE}
#install.packages("lubridate")
library(lubridate)
```

```{r}
age <- today() - ymd("2000-07-18")
age
```

```{r}
time_length(age, "year")
```


convert our date data into a lubridate format

```{r}
vax$as_of_date <- ymd(vax$as_of_date)
```

how many days since the first vaccination reported in the data set?

```{r}
today() - vax$as_of_date[1]
```

> Q9. How many days have passed since the last update of the dataset? 

```{r}
today() - vax$as_of_date[nrow(vax)]
```

> Q10. How many unique dates are in the dataset?

```{r}
length(unique(vax$as_of_date))
```


> Working with zip codes

use the zipcodeR package

```{r, message=FALSE}
#install.packages("zipcodeR")
library(zipcodeR)
```

locate la jolla

```{r}
geocode_zip('92037')
```

calculate the distance between two zipcodes

```{r}
zip_distance('92037','91362')
```

use reverse_zipcode() to pull census data

> Focus on San Diego Area

restrict to vax$county == "San Diego" entries

using r

```{r}
# Subset to San Diego county only areas
sd <- vax[2,]
```

```{r}
library(dplyr)

sd <- filter(vax, county == "San Diego")

nrow(sd)
```

> Q11. How many distinct zip codes are listed for San Diego County?

```{r}
length(unique(sd$zip_code_tabulation_area))
```


> Q12. What San Diego County Zip code area has the largest 12 + Population in this dataset?

```{r}
pop <- which.max(sd$age12_plus_population)
sd$zip_code_tabulation_area[pop]
```

Using dplyr select all San Diego “county” entries on “as_of_date” “2022-02-22” 

```{r}
feb <- filter(sd, as_of_date=="2022-02-22")
```


> Q13. What is the overall average “Percent of Population Fully Vaccinated” value for all San Diego “County” as of “2022-02-22”? 

```{r}
sd$as_of_date[nrow(sd)]
```

let's work with the most recent data

```{r}
latest <- filter(sd, as_of_date=="2022-03-01")
mean(latest$percent_of_population_fully_vaccinated, na.rm=TRUE)
```


> Q14. Using either ggplot or base R graphics make a summary figure that shows the distribution of Percent of Population Fully Vaccinated values as of “2022-03-01”?

```{r, message=FALSE}
library(ggplot2)
```

```{r}
ggplot(latest)+aes(percent_of_population_fully_vaccinated)+geom_histogram()
```

> Focus on UCSD/La Jolla

```{r}
ucsd <- filter(sd, zip_code_tabulation_area=="92037")
ucsd[1,]$age5_plus_population
```

> Q15. Using ggplot make a graph of the vaccination rate time course for the 92037 ZIP code area

```{r}
baseplot <- ggplot(ucsd) +
  aes(x=as_of_date,
      y=percent_of_population_fully_vaccinated) +
  geom_point()+
  geom_line() +
  geom_line(group=1) +
  ylim(c(0,1)) +
  labs(x="Date", y="Percent Vaccinated", title="Vaccination rate for La Jolla 92037")
baseplot
```

> Comparing to similar sized areas

```{r}
# Subset to all CA areas with a population as large as 92037
vax.36 <- filter(vax, age5_plus_population > 36144 &
                as_of_date == "2022-03-01")

#head(vax.36)
```


> Q16. Calculate the mean “Percent of Population Fully Vaccinated” for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2022-03-01”. Add this as a straight horizontal line to your plot from above with the geom_hline() function?

```{r}
ave.36 <- mean(vax.36$percent_of_population_fully_vaccinated, na.rm=TRUE)
ave.36
```

```{r}
baseplot+geom_hline(yintercept=ave.36)
```


> Q17. What is the 6 number summary (Min, 1st Qu., Median, Mean, 3rd Qu., and Max) of the “Percent of Population Fully Vaccinated” values for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2022-03-01”?

```{r}
summary(vax.36$percent_of_population_fully_vaccinated)
```


> Q18. Using ggplot generate a histogram of this data.

```{r}
ggplot(vax.36) +
  aes(percent_of_population_fully_vaccinated) +
  geom_histogram() +
  xlim(c(0,1))
```

> Q19. Is the 92109 and 92040 ZIP code areas above or below the average value you calculated for all these above?

```{r}
vax %>% filter(as_of_date == "2022-03-01") %>%  
  filter(zip_code_tabulation_area=="92040") %>%
  select(percent_of_population_fully_vaccinated)

vax %>% filter(as_of_date == "2022-03-01") %>%  
  filter(zip_code_tabulation_area=="92109") %>%
  select(percent_of_population_fully_vaccinated)
```

both areas have a lower fully vaccinated average

> Q20. Finally make a time course plot of vaccination progress for all areas in the full dataset with a age5_plus_population > 36144

```{r}
vax.36.all <- filter(vax, age5_plus_population > 36144)


ggplot(vax.36.all) +
  aes(as_of_date,
      percent_of_population_fully_vaccinated, 
      group=zip_code_tabulation_area) +
  geom_line(alpha=0.2, color="blue") +
  ylim(c(0,1)) +
  labs(x="Date", y="Percent Vaccinated",
       title="Vaccination rate across California",
       subtitle="Only areas with a population above 36K are shown") +
  geom_hline(yintercept = ave.36, linetype=2)
```
> 21. How do you feel about traveling for Spring Break and meeting for in-person class afterwards?

I think it is okay as long as everyone is testing after they come back, but I doubt that most people would do that so it makes me question if we should go back in person immediately. However, I don't think classes should be online anymore, that if everyone is wearing masks inside it should be fine.








